---
title: "Storm Data Analysis"
author: "Wenhao Y"
date: "April 11, 2016"
output: md_document
---


Install Packages
```{r}
library(R.utils)
library(assertthat)
library(dplyr)
library(ggplot2)
library(xtable)
```

Data Processing
```{r}
#Setting WD
setwd("C:/Users/sharon/Desktop/")

#Unzip and read .csv file into the variable data
unzip <- bzfile("repdata-data-StormData.csv.bz2", "r")
data <- read.csv(unzip, stringsAsFactors = FALSE)
close(unzip)


```


Subsetting data and creating new variable

```{r}
x <- which(colnames(data) %in% c("BGN_DATE", "PROPDMG", "CROPDMG", "EVTYPE", 
    "INJURIES", "FATALITIES"))
data <- data[, x]
head(data)

#Formatting date and time
data$YEAR <- as.integer(format(as.Date(data$BGN_DATE, "%m/%d/%Y 0:00:00"), "%Y"))
head(data)

#To uppercase
data$EVTYPE <- toupper(data$EVTYPE)
head(data)

# creates new variable
data$ECONOMICDMG <- data$PROPDMG + data$CROPDMG
head(data)

# Select only positive value data
data <- subset(data, data$FATALITIES > 0 | data$ECONOMICDMG > 0 | data$INJURIES > 
    0)
head(data)
```

Aggregating Data
```{r}
library(plyr)

# data aggregated by YEAR & EVTYPE.
#ddply -> For each subset of a data frame, apply function then combine results into a data frame.

eventYear <- ddply(data[, -1], .(YEAR, EVTYPE),
                   .fun = function(x) {
                         return(
                           c(sum(x$FATALITIES), sum(x$ECONOMICDMG), sum(x$INJURIES))
                              )
                                      }
                   )
names(eventYear) <- c("YEAR", "EVTYPE", "FATALITIES", "ECONOMICDMG", "INJURIES")
head(eventYear)
```

Events are grouped by categories
```{r}
#Function that calculates the events by categories (13 categories described in the synopsis)

#grepl -> search for matches to argument pattern within each element of a character vector

eventCategory <- function(x) {
    ev <- x$EVTYPE[1]
    if (grepl("LIG(H|N)T(N|)ING|TORNADO|T(H|)U(N|)(DER|ER|DEER|DERE)(STORM|STROM|TORM)|TSTM|HAIL", 
        ev)) {
        category <- "Convection"
    } else if (grepl("WINT(ER|RY)|ICE|AVALANC(H|)E|SNOW|BLIZZARD|FREEZ|ICY|FROST", 
        ev)) {
        category <- "Winter"
    } else if (grepl("COLD|HEAT|HOT|TEMPERATURE|COOL|WARM", ev)) {
        category <- "Extreme Temp"
    } else if (grepl("FLOOD| FLD$", ev)) {
        category <- "Flood"
    } else if (grepl("COASTAL|TSUNAMI|RIP CURRENT|MARINE|WATERSPOUT|SURF|SLEET|SEAS|(HIGH|RISING|HEAVY) (WAVES|SWELLS|WATER)", 
        ev)) {
        category <- "Marine"
    } else if (grepl("TROPICAL|HURRICANE|STORM SURGE|TYPHOON", ev)) {
        category <- "Tropical Cyclones"
    } else if (grepl("WIND|MICROBURST", ev)) {
        category <- "High Wind"
    } else if (grepl("FIRE", ev)) {
        category <- "Fire"
    } else if (grepl("RAIN|PRECIP", ev)) {
        category <- "Rain"
    } else if (grepl("DROUGHT|DUST", ev)) {
        category <- "Drought/Dust"
    } else if (grepl("LANDSLIDE|MUD.*SLIDE", ev)) {
        category <- "Landslide"
    } else if (grepl("FOG|VOG", ev)) {
        category <- "Fog"
    } else {
        category <- "Others"
    }

    x$EVGROUP <- rep(category, dim(x)[1])
    return(x)
}
eventYear <- ddply(eventYear, .(EVTYPE), .fun = eventCategory)
head(eventYear)

#We organize the data to show FATALITIES, ECONOMICDMG and INJURIES
#by YEAR and EVGROUP

groupYear <- ddply(eventYear, .(YEAR, EVGROUP), .fun = function(x) {
    return(c(sum(x$FATALITIES), sum(x$ECONOMICDMG), sum(x$INJURIES)))
})

names(groupYear) <- c("YEAR", "EVGROUP", "FATALITIES", "ECONOMICDMG", "INJURIES")
head(groupYear)

# calculate average annual damage by group
eventFirstYear <- ddply(groupYear, .(EVGROUP), .fun = function(x) {
    return(c(min(x$YEAR)))
})
names(eventFirstYear) <- c("Weather.Event", "First.Year")
head(eventFirstYear)
```


Notice that events start at different year.  To even this out, we start the group at 1993. 
```{r}
## start data analysis at 1993
groupYear <- subset(groupYear, YEAR >= 1993)

# calculate average annual damage by group
byGroup <- ddply(groupYear, .(EVGROUP), .fun = function(x) {
    return(c(mean(x$FATALITIES), mean(x$ECONOMICDMG), mean(x$INJURIES)))
})
names(byGroup) <- c("EVGROUP", "AVG.FATALITIES", "AVG.ECONOMICDMG", "AVG.INJURIES")
head(byGroup)
```

RESULTS

Results section 1 - Health

This histograms Show fatalities and injuries for weather events.
```{r}
# Graph libraries
library(scales)

# average annual populational damage by group of event
byGroup$EVGROUP <- with(byGroup, reorder(EVGROUP, -AVG.FATALITIES))
g <- ggplot(byGroup, aes(x = EVGROUP))
g + geom_bar(aes(weight = AVG.FATALITIES, fill = AVG.FATALITIES, stat="count"), binwidth = 5, 
    color = "black") + ggtitle("Average Fatalities") + ylab("# Fatalities") + 
    xlab("Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# average annual populational damage by group of event

byGroup$EVGROUP <- with(byGroup, reorder(EVGROUP, -AVG.INJURIES))
g <- ggplot(byGroup, aes(x = EVGROUP))
g + geom_bar(aes(weight = AVG.INJURIES, fill = AVG.INJURIES, stat="count")+
    color = "black") + ggtitle("Average Injuries") + ylab("# Injuries") + xlab("Weather Event") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Results section 2 - Economic damage

Histogram of weather event harm to the economy.
```{r}
# average annual economical damage by group of event
byGroup$EVGROUP <- with(byGroup, reorder(EVGROUP, -AVG.ECONOMICDMG))
g <- ggplot(byGroup, aes(x = EVGROUP))
g + geom_bar(aes(weight = AVG.ECONOMICDMG, fill = AVG.ECONOMICDMG, stat="count")+
    color = "black") + ggtitle("Average Economic Damage") + ylab("Economic damage") + 
    xlab("Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Conclusion

According to this analysis, we can tell that "Extreme temperatures" and "Convection" have the highest fatalities.  Similarly, "Convection" and "Extreme temperatures" are respectively also the highest ranked in "Average Injuries".

In the scope of economic damage, "convection" and "flood" ranks the highest.